---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\maketitle
# Introduction

## Outline
\tableofcontents

# Sequential Execution

## Sequential Execution

- Run a script
  - step by step;
  - from start to end (or other condition).
```{r, eval = FALSE}
A <- "Well"
B <- "hello there."
paste(A, B, sep = ", ")
rm(A, B, C) # Explain the error.
```

# Iterative Execution

## A.K.A. Looping

- Execute statement(s) repeatedly
  a. over a set of values
  b. as long as some condition holds
  c. until an abort condition is met
- Includes: for, while, and repeat
- Typical use-case: transform several variables

## for()-Loops[^1]

- repeats statements for each element on an input set

```{r, eval = FALSE}
# Generic example
for (VALUE in THAT) { # Do THIS for each VALUE in THAT
  THIS
}
# A first working example
for (value in c("Waiting", "for", "statistics.")) {
  print(value)
}
```

- for() creates an object called VALUE
- reassigns VALUE for each element in the set THAT

[^1]: People don't like for(). For alternatives see https://bit.ly/2IEbeGj.

## for()-Loops, contd.

- for() returns nothing unless told to[^2]
- Save the output to an object
- Good practice:
  - Execute on a set of integers
  - Index both object and storage simultaneously

```{r, eval = FALSE}
words <- c("So", "how's", "looping", "so", "far?")
chr <- vector("character", length = length(words))
for (i in 1:length(words)){
	chr[i] <- words[i]
}
```

[^2]: "for loops are like Las Vegas: what happens in a for loop stays in a for loop" (Gorrelmund 2014: 164).

## Quick Exercise

Remember last session's data management challenge? Let's
try to express our solution as a for()-Loop.
```{r, eval = FALSE}
grade_quantiles <- quantile(
  student_data[, "grade"], probs = c(.2,.4,.6,.8)
)
student_data[, "grade_alp"] <- "F"
student_data[
  student_data[, "grade"] > grade_quantiles["20%"],
  "grade_alp"
] <- "D"
# ... and so on until A.
```

## while()-Statements
- Rerun statement(s) as long as some condition is TRUE
- Remember ["Groundhog Day"](https://www.imdb.com/title/tt0107048/)?

```{r, eval = FALSE}
k <- 0
while (k < 20) {  # statement evaluates to ONE value
  cat(k, "and counting...", "\n")
  k <- k + 1  # Make sure that your condition fails!
}
```

## repeat()-Statements

- Rerun statement(s) until **break** statement is met

```{r, eval = FALSE}
chr <- "All work and no play makes Jack a dull boy"
k <- 0
repeat {
	print(chr)
	k <- k + 1
	if (k > 100) break
}
```

\begin{block}{But\dots}
  How do we tell $\mathcal{R}$ to execute some code conditionally?
\end{block}

# Conditional Execution

## Intuition
  - How does the absolute value $|x|$ work algorithmically?

\begin{figure}
  \centering
  \includegraphics[scale = .6]{./fig/abs_value.png}
\end{figure}

- *Different* operations follow depending on some condition
- $\rightarrow$ code handles parallel cases


## Detour: Logical Tests in *R*

- Check the truth value of some condition
- All tests return a logical values
- Obey element-wise execution and recycling

\begin{table}
  \centering
  \begin{tabular}{lll}
      \toprule
      Operator & Syntax & Test\\
      \midrule
      $>$ & a $>$ b & Is a greater than b?\\
      $>=$ & a $>=$ b & Is a at least equal to b?\\
      $<$ & a $<$ b & Is a less than b?\\
      $<=$ & a $<=$ b & Is a at most equal to b?\\
      $==$ & a $==$ b & Is a equal to b?\\
      $!=$ & a $!=$ b & Is a not equal to b?\\
      \%in\% & a \%in\% c(a, b, c) & Is a in the set c(a, b, c)?\\
      \bottomrule
  \end{tabular}
\end{table}


## Longer Detour: Boolean Operators in *R*

- Operate on several logical tests
- *Usually* return a single logical value

\begin{table}
    \centering
    \begin{tabular}{lll}
        \toprule
        Operator & Syntax & Test\\
        \midrule
        \& & cond1 \& cond2 & Are both conditions true?\\
        $|$ & cond1 \& cond2 & Is at least one condition true?\\
        xor & xor(cond1, cond2) & Is either cond1 or cond2 true?\\
        ! & !cond1 & Negate cond1. \\
        any() & any(cond1, cond2) & Is any condition true?\\
        all() & all(cond1, cond2) & Are all conditions true?\\
        \%in\% & a \%in\% c(a, b, c) & Is a in the set c(a, b, c)?\\
        \bottomrule
    \end{tabular}
\end{table}


## if()-Statements

- Code executes if and only if some condition is TRUE
- Condition should evaluate to a single TRUE/FALSE statement

```{r, eval = FALSE}
if (THIS) { # If this is TRUE
  THAT # then do THAT.
}
x <- -4
if (x < 0) {
  x <- -1 * x
}
x
```

## if()-Statements: What will this return?

```{r, eval = FALSE}
# Example 1 =========================================
x <- 1
if (TRUE) {
  x <- 2
}
# Example 2 =========================================
x <- 1
if (x == 1) {
  x <- 2
  if (x == 1) {
    x <- 3
  }
}
```

## else()-Statements

- tell $\mathcal{R}$ what to do should if() evaluate to FALSE
- multiple if/else statements can be nested

```{r, eval = FALSE}
if (this) {
  Plan A
} else {
  Plan B
}
dec <- pi # Example: Round a decimal to integer
if (dec - trunc(dec) >= 0.5) {
  dec <- trunc(dec) + 1
} else {
  dec <- trunc(dec)
}
```

## But I want to operate on an entire vector!

- Looping: Use vectorized functions, e.g. apply(), sapply(), etc.
- Conditional execution: Use `ifelse(<test>, <yes>, <no>)`

```{r}
x <- 1:6
evens <- ifelse(x %% 2 == 0, 1, 0)
rbind(x, evens)
```

# Summary

## What was that about?

- Code can be executed sequentially, iteratively, or conditionally
- Sequential execution is the norm
- Iterative execution runs the same code repeatedly
  - **for()** reruns statement(s) for all members of a set
  - **while()** reruns statement(s) as long as a condition is met
  - **repeat()** reruns statement(s) until it encounters **break**
- Conditional execution manages parallel cases
  - **if()** runs statesment(s) if a condition evaluates to TRUE
  - **else()** runs statesment(s) if that same condition is FALSE
  - **ifelse()** is a vectorized version

# Practical Challenge

## In a world where humanity forgot the value of $\pi$ \dots

\begin{columns}
  \begin{column}{.5\textwidth}
    {\dots} we will uplift civilization by Monte Carlo simulation.
    \bigskip

    Set up a simulation which allows you to generate an estimate
    of $\pi$ from the chance to hit a circle perfectly inscribed
    in a square with a randomly thrown dart.
    \bigskip

    You can do this in four lines!
  \end{column}
  \begin{column}{.5\textwidth}
    \begin{figure}
      \centering
      \includegraphics[scale = .5]{./fig/circle-in-square-red.png}
    \end{figure}
    \begin{align*}
      p(Hit) &= \frac{A_{ci}}{A_{sq}} = \frac{\pi r^2}{(2r)^2}\\
      \pi &= 4p(Hit)
    \end{align*}
  \end{column}
\end{columns}