---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---

\maketitle


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
asoiaf <- read.csv2(
  "~/github/teaching/pcqr/r_intro/06/dta/asoiaf.csv",
  stringsAsFactors=FALSE
)
```


# Introduction


## Before we start\dots

- Quit \& reopen RStudio.
- Open the notebook "notebook_06.Rmd" in RStudio
- Load "./06/dta/asoiaf.csv" from the course material.
- Execute the following code below. What does it do?

```{r, eval = TRUE}
asoiaf[, "is_dead"] <- !is.na(asoiaf[, "book_of_death"])
```


## What do we intent to do?

- **Question**: What's the chance that Jon Snow is going to die?
- **Means**: Regression on a linear combination of predictors
$$p(Death = 1 | x, \beta)  = \sum_K \beta_kx_k$$
- **Caveat**: Chance of death is not a well-behaved response.


## Outline
\tableofcontents


# Linear Models in $\mathcal{R}$


## Generic Format of Fitting GLMs

```{r, eval = FALSE}
fit <- lm(
  formula = <formula>,
  family = <family>(link = "<link>"),
  data = <data>,
  subset = <subset>,
  weights = <weights>, # Be careful! Meaning changes
                       # depending on <family>.
  na.action = na.omit, # Retains only complete cases.
  <...> # Further options.
)
```

## $\mathcal{R}$'s Formula Interface\footnote{
  Adapted from Kabacoff, R. 2011. R in Action. Shelter Island: Manning Publications, p. 178.
}

**Generic Example**
$$ y \sim x_1 + x_2 + \dots + x_k $$

**Formula Creation**

\resizebox{\textwidth}{!}{
  \input{./tbl/formula_arguments}
}

## $\mathcal{R}$'s Formula Interface, contd.

**Exercise** How would you write the following formulas?[^1]

1. $y = a + x + z + xz$
2. $y = a + x + x^2 + x^3$
3. $log_{10}(y) = x + z + w + xz + xw + wz$
4. $y$ as a function of variables in the data but $k$

[^1]: Assume $a$ is the constant.

## Get Your Hands Dirty

Now it's your turn. Use the **asoiaf** data to

- regress **is_dead** on
- **allegiances**,
- the full interaction of **gender** and **nobility**, and
- a cubic polynomial on **age_in_chapters**.
- Save the results to an object called **myfit**.

## One Solution to the Exercise

```{r, results = FALSE}
myfit <- glm(
  formula = is_dead ~ 0 + allegiances +
    gender * nobility +
    age_in_chapters + I(age_in_chapters^2) +
    I(age_in_chapters^3),
  data = asoiaf
)
```

# Working With Results

## A Menu of Typical Options\footnote{
  Adapted from Kabacoff, R. 2011. R in Action. Shelter Island:
  Manning Publications, p. 179.
}

\begin{table}
  \centering
  \input{./tbl/accessor_functions}
\end{table}

## How to Predict New Data

**Generic Sequence**

1. Define scenarios to predict
2. Create a date frame which contains those scenarios
3. Use predict() to return quantities of interest
4. Summarize the results


## Example: How does the chance to die change over time?

```{r, eval = FALSE}
# Steps 1 & 2 Prediction scenarios
pred_dta <- data.frame(
  allegiances = "Baratheon",
  gender = mean(asoiaf$gender),
  nobility = mean(asoiaf$nobility),
  age_in_chapters = 1:343, stringsAsFactors = FALSE
)
# Step 3 Predict
pred_dta[, "fitted"] <- predict(myfit,
  newdata = pred_dta)
# Step 4 Summarize
ggplot(data = pred_dta,
  aes(x = age_in_chapters, y = fitted)) + geom_line()
```


# Generalized Linear Models


## Intuition

- **Problem**
  a. We don't obseverve probabilities but discrete events.
  b. Probabilities are restricted to $[0, 1]$, but $\bf{X\beta}$
    can take any value.
- **Challenge**: Map $\bf{X\beta}$ to the domain of our response
- Applies to many quantities of interest, e.g.,
  - Household income
  - Satisfaction with democracy
  - Number of bills per session of parliament
  - \dots
- **GLMs**: link function $g(\cdot)$ relates response to $\bf{X\beta}$
  - logit transformation [$ln(\frac{p}{1 - p})$] for binary DVs
  - natural log ($ln(\mu)$) for count data

## Generic Format of Fitting GLMs

```{r, eval = FALSE}
fit <- glm(
  formula = <formula>,
  family = <family>(link = "<link>"),
  data = <data>,
  weights = <weights>, # Be careful! Meaning changes
                       # depending on <family>.
  subset = <subset>,
  na.action = na.omit, # Retains only complete cases.
  <...> # Options to tweak the optimizer.
)
```


## Family Generators and Link Functions in glm()\footnote{
  Adapted from Fox, J. and S. Weisberg. 2011. An R Companion to
  Applied Regression. 2nd ed. London: SAGE, pp. 231, 233.
}

**A Practical Example**

```{r, eval = FALSE}
glm(<...>, family = binomial(link = "logit"), <...>)
```

\resizebox{\textwidth}{!}{
  \input{./tbl/link_functions}
}

## Get Your Hands Dirty

Now it's your turn. Is John Snow going to die? Setup
possible scenarios and evaluate the results.


## One Possible Solution
```{r, eval = FALSE}
jon_snow <- which(asoiaf$name == "Jon Snow")
pred_dta <- asoiaf[rep(jon_snow, 3), ]; rm(jon_snow)
pred_dta[2, "allegiances"] <- "Stark"
pred_dta[3, "allegiances"] <- "Targaryen"
pred_dta[, "fitted"] <- predict(
  myfit, newdata = pred_dta, type = "response"
)
pred_dta[, "fitted"]
```

# Checking Assumptions

## Checking Assumptions

- Always check your diagnostic plots
```{r, eval = FALSE}
plot(myfit)
```

- For detailed instructions see: Fox, J. and S. Weisberg. 2011.
  An R Companion to Applied Regression. 2nd ed. London: SAGE.

# Summary

## Summary

- base $\mathcal{R}$ offers many probability models
- Numerous extensions are available (see the [CRAN Taskviews](https://cran.r-project.org/web/views/))
- Discussion of marginal effects requires some acrobatics
  + Familiarize yourselves with Fox's and Weisberg's [car](https://cran.r-project.org/web/packages/car/index.html) package
  + Check out Daniel Lüdecke's [ggeffects](https://cran.r-project.org/web/packages/ggeffects/index.html) package
