---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
asoiaf <- read.csv2(
  "~/github/teaching/pcqr/r_intro/05/dta/asoiaf.csv",
  stringsAsFactors=FALSE
)
```

\maketitle

# Introduction

## Before we start\dots

- Quit & reopen RStudio.
- Load "./06/dta/asoiaf.csv"" from the course material.
  - **Remember**: Uncheck the option "Strings as factors"
- Open a new script file.
- Execute the following code:
```{r, eval = FALSE}
asoiaf[, "died"] <- !is.na(asoiaf[, "book_of_death"])
```
- Install the "car" package.

## What do we intent to do?

- **Question**: What's the chance that Jon Snow is going to die?
- **Means**: Regression on a linear combination of predictors
$$p(Death = 1 | \bf{X}, \bf{\beta})  = \beta_0 + \sum_K \beta_kx_k$$
- **Problem**: Chance of death is not a well-behaved response.
  a. We don't obseverve probabilities but discrete events.
  b. Probabilities are restricted to $[0, 1]$, but $\bf{X\beta}$
    can take any value.
- **Challenge**: Map the linear combination $\bf{X\beta}$ into a
  domain which fits our response.

## Some Intuition on GLMs

- Applies to many quantities of interest, e.g.,
  - Household income
  - Satisfaction with democracy
  - Number of bills per session of parliament
  - \dots
- 




## Outline
\tableofcontents

# The Basics of Running GLMs in $\mathcal{R}$

## Generic Format of Fitting GLMs

```{r, eval = FALSE}
fit <- glm(
  formula = <formula>,
  family = <family>(link = "<link>"),
  # Defaults to gaussian(link = "identity"). Therefore
  # we skip the lm() function and OLS.
  data = <data>,
  weights = <weights>, # Be careful! Meaning changes
                       # depending on <family>.
  subset = <subset>,
  na.action = na.omit, # Retains only complete cases.
  <...> # Options to tweak the optimizer.
)
```

## $\mathcal{R}$'s Formula Interface\footnote{
  Adapted from Kabacoff, R. 2011. R in Action. Shelter Island: Manning Publications, p. 178.
}

**Generic Example**
$$ y \sim x_1 + x_2 + \dots + x_k $$
**Formula Creation**

\resizebox{\textwidth}{!}{
  \input{./tbl/formula_arguments}
}
 
## $\mathcal{R}$'s Formula Interface, contd.

**Exercise** How would you write the following formulas?

1. $y = x + z + xz$
2. $y = x + x^2 + x^3$
3. $log_e(y) = x + z + w + xz + xw + wz$
4. $y$ as a function of variables in the data but $k$
  
## Family Generators and Link Functions in glm()\footnote{
  Adapted from Fox, J. and S. Weisberg. 2011. An R Companion to
  Applied Regression. 2nd ed. London: SAGE, pp. 231, 233.
}

**A Practical Example**

```{r, eval = FALSE}
glm(<...>, family = binomial(link = "logit"), <...>)
```

\resizebox{\textwidth}{!}{
  \input{./tbl/link_functions}
}

## Get Your Hands Dirty

Now it's your turn. Use the **asoiaf** data to

- regress **died** on
- **allegiances**,
- the full interaction of **gender** and **nobility**, and
- a cubic polynomial on **age_in_chapters**.
- This should be a **logistic** regression model.
- Save the results to an object called **myfit**.

# Working With Regression Results

# Testing Assumptions

