---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\maketitle
# Introduction

## Outline
\tableofcontents



# Sequential Execution

## Sequential Execution

- Run a script
  - step by step;
  - from start to end.
```{r, eval = FALSE}
A <- "Well"
B <- "hello there."
paste(A, B, sep = ", ")
rm(A B) # Explain the error.
```

# Iterative Execution

## A.k.a. Looping

- Execute statement(s) repeatedly
  a. over a set of values
  b. as long as some condition holds
  c. until an abort condition is met
- Includes: for, while, and repeat
- Typical use-case: transform several variables

## for()-Loops[^1]
- repeats statements for each element on an input set
```{r, eval = FALSE}
# Generic example
for (VALUE in THAT) { # Do THIS for each VALUE in THAT
  THIS
}
# A first working example
for (value in c("Waiting", "for", "statistics.")) {
  print(value)
}
```
- for() creates an object called VALUE
- reassigns VALUE for each element in the set THIS

[^1]: People don't like for(). For alternatives see https://bit.ly/2IEbeGj.

## for()-Loops, contd.
- for() returns nothing unless told to[^2]
- Save the output to an object
- Good practice:
  - Execute on a set of integers
  - Index both object and storage simultanously
```{r, eval = FALSE}
words <- c("So", "how's", "looping", "so", "far?")
chr <- vector("character", length = length(words))
for (i in 1:length(words)){
	chr[i] <- words[i]
}
```

[^2]: "for loops are like Las Vegas: what happens in a for loop stays in a for loop" (Gorrelmund 2014: 164).

## While()-Loops
- Rerun statement(s) as long as some condition is TRUE
- Condition should be a logical test
- Remember ["Groundhog Day"](https://www.imdb.com/title/tt0107048/)
  - Include a change of condition in the while()'s body!
```{r, eval = FALSE}
k <- 0
while (k < 20) {
  k <- k + 1 
  print("Still running") 
}
```
- Returns anything unless told to

## Repeat()-Statements
- Reruns statement(s) until meets **break**
```{r, eval = FALSE}
chr <- "All work and no play makes Jack a dull boy" 
k <- 0
repeat {
	print(chr)
	k <- k + 1
	if (k > 100) break
}
```

\begin{block}{But\dots}
  How do we tell $\mathcal{R}$ to execute some code conditionally?
\end{block}

# Conditional Execution

## if()-Statements

## else()-Statements

# Practical Challenges

## Imagine a world without $\pi$

## Simulate collider bias



# Summary
