---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\maketitle
# Introduction

## Outline
\tableofcontents

# Sequential Execution

## Sequential Execution

- Run a script
  - step by step;
  - from start to end.
```{r, eval = FALSE}
A <- "Well"
B <- "hello there."
paste(A, B, sep = ", ")
rm(A, B, C) # Explain the error.
```

# Iterative Execution

## A.K.A. Looping

- Execute statement(s) repeatedly
  a. over a set of values
  b. as long as some condition holds
  c. until an abort condition is met
- Includes: for, while, and repeat
- Typical use-case: transform several variables

## for()-Loops[^1]
- repeats statements for each element on an input set
```{r, eval = FALSE}
# Generic example
for (VALUE in THAT) { # Do THIS for each VALUE in THAT
  THIS
}
# A first working example
for (value in c("Waiting", "for", "statistics.")) {
  print(value)
}
```
- for() creates an object called VALUE
- reassigns VALUE for each element in the set THIS

[^1]: People don't like for(). For alternatives see https://bit.ly/2IEbeGj.

## for()-Loops, contd.
- for() returns nothing unless told to[^2]
- Save the output to an object
- Good practice:
  - Execute on a set of integers
  - Index both object and storage simultanously
```{r, eval = FALSE}
words <- c("So", "how's", "looping", "so", "far?")
chr <- vector("character", length = length(words))
for (i in 1:length(words)){
	chr[i] <- words[i]
}
```

[^2]: "for loops are like Las Vegas: what happens in a for loop stays in a for loop" (Gorrelmund 2014: 164).

## Quick Exercise

Remember last session's data management challenge? Let's
try to express our solution as a for()-Loop.
```{r, eval = FALSE}
grade_quantiles <- quantile(
  student_data[, "grade"], probs = c(.2,.4,.6,.8)
)
student_data[, "grade_alp"] <- "F"
student_data[
  student_data[, "grade"] > grade_quantiles["20%"],
  "grade_alp"
] <- "D"
# ... and so on until A.
```



## While()-Loops
- Rerun statement(s) as long as some condition is TRUE
- Condition should be a logical test
- Remember ["Groundhog Day"](https://www.imdb.com/title/tt0107048/)
  - Include a change of condition in the while()'s body!
```{r, eval = FALSE}
k <- 0
while (k < 20) {
  k <- k + 1 
  print("Still running") 
}
```
- Returns nothing unless told to

## Repeat()-Statements
- Reruns statement(s) until meets **break**
```{r, eval = FALSE}
chr <- "All work and no play makes Jack a dull boy" 
k <- 0
repeat {
	print(chr)
	k <- k + 1
	if (k > 100) break
}
```

\begin{block}{But\dots}
  How do we tell $\mathcal{R}$ to execute some code conditionally?
\end{block}

# Conditional Execution

## Intuition
  - How does the absolute value $|x|$ algorithmically work?
  
\begin{figure}
  \centering
  \includegraphics[scale = .6]{./fig/abs_value.png}
\end{figure}

- *Different* operations follow depending on some condition
- $\rightarrow$ code handles parallel cases

## if()-Statements

- Code executes if and only if some condition is TRUE
- Condition **must** evaluate to a single TRUE/FALSE statement
```{r, eval = FALSE}
if (THIS) { # If this is TRUE
  THAT # then do THAT.
}
x <- 4
if (x < 0) {
  x <- -1 * x
}
```

## if()-Statements: What will this return?
```{r, eval = FALSE}
# Example 1 =========================================
x <- 1
if (TRUE) {
  x <- 2
}
# Example 2 =========================================
x <- 1
if (x == 1) {
  x <- 2
  if (x == 1) {
    x <- 3
  }
}
```

## else()-Statements
- tell $\mathcal{R}$ what to do should if() evaluate to FALSE
- multiple if/else statements can be nested
```{r, eval = FALSE}
if (this) {
  Plan A
} else {
  Plan B
}
dec <- 3.141 # Example: Round a decimal to integer
if (dec - trunc(dec) >= 0.5) {
  dec <- trunc(dec) + 1
} else {
  dec <- trunc(dec)
}
```

# Summary

## What was that about?

- Code can be executed sequentially, iteratively, or conditionally
- Sequential execution is the norm
- Iterative execution runs the same code repeatedly
  - **for()** reruns statement(s) for all members of a set
  - **while()** reruns statement(s) as long as a condition is met
  - **repeat()** reruns statement(s) until it encounters **break**
- Conditional execution manages parallel cases
  - **if()** runs statesment(s) if a condition evaluates to TRUE
  - **else()** runs statesment(s) if that same condition is FALSE

# Practical Challenges

## In a world where humanity forgot the value of $\pi$ \dots

\begin{columns}
  \begin{column}{.5\textwidth}
    {\dots} we will uplift civilization by Monte Carlo simulation.
    \bigskip
    
    Set up a simulation which allows you to generate an estimate
    of $\pi$ from the chance to hit a circle perfectly inscribed
    in a square with a randomly thrown dart.
  \end{column}
  \begin{column}{.5\textwidth}
    \begin{figure}
      \centering
      \includegraphics[scale = .5]{./fig/circle-in-square-red.png}
    \end{figure}
    \begin{align*}
      p(Hit) &= \frac{A_{ci}}{A_{sq}} = \frac{\pi r^2}{(2r)^2}\\
      \pi &= 4p(Hit)
    \end{align*}
  \end{column}
\end{columns}

## Understanding Collider Bias

In causal inference, collider bias results when we condition on a 
variable that is causally influenced by one or more other 
variables. The effect will be a spurious correlation between the
collider's ancestors. Setup a simulation to demonstrate collider
bias.

\begin{figure}
  \centering
  \includegraphics[scale = .5]{./fig/collider_bias.png}
  \caption{Collider Bias Visualized\footnote{
    Ellwert, F. 2013. Graphical Causal Models. In: S.L. Morgan 
      (ed.), Handbook of Causal Analysis for Social Research, p. 
      251
    }
  }
\end{figure}
