---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
asoiaf <- read.csv(
  "~/github/teaching/pcqr/r_intro/05/dta/asoiaf.csv",
  sep=";",
  stringsAsFactors=FALSE
)
```

\maketitle

# Introduction

## So far, you have learned...

>- to navigate the $\mathcal{R}$ environment,
>- to create, index, and modify objects,
>- to take advantage of control flow statements,
>- generate decent looking graphs.
>- You are now ready to unleash $\mathcal{R}$ on real data.

## What are we going to do?

We are going to plunge head first into data analysis, taking
advantage of a data set which includes all *named*
characters from George R.R. Martin's "A Song of Ice and
Fire".[^1] In the end, you will be able to offer much insight on
the mother of all cocktail party questions: **Is Jon Snow going to die?**

To get started:

1. Quit & reopen $\mathcal{R}$.
2. Load "./05/dta/asoiaf.csv" from the course material.
    - **Note**: Uncheck the option "Strings as factors".

[^1]: O'Neill, M. 2016. Game of Thrones. https://bit.ly/2qjUfQ2 (last access: 10/08/2018).

## Outline
\tableofcontents

# Descriptive Statistics

## Glancing at the data
Enter each of the following commands. Explain the output.
```{r, eval = FALSE}
str(asoiaf); summary(asoiaf)
mean(asoiaf[, "book_intro_chapter"], na.rm = TRUE)
sd(asoiaf[, "book_intro_chapter"], na.rm = TRUE)
by(
  data = asoiaf[, 6:7],
  INDICES = asoiaf[, c("gender")], FUN = summary
)
```

>- What have we learned?
>- str() returns a brief description of the object
>- base::summary() returns the five-point-summary and mean
>- mean() & sd() return just that, require na.rm = TRUE
>- by() may create summary statistics by group

# Contingency Tables

## Simple N-way Contingency Tables

- table() creates N-way contingency tables
```{r, eval = FALSE}
table(asoiaf[, "death_year"]) # 1way
table(asoiaf[, "death_year"], asoiaf[, "gender"]) # 2way
table(asoiaf[, "death_year"], asoiaf[, "gender"], asoiaf[, "nobility"]) # 3way
```

Do you notice anything about the output?
Check summary(asoiaf[, "death_year"]) if necessary.

## Test

# Adventures in Covariance

# Simple Hypotheses Tests

# Summary
