--- 
title: "Crisis of Democracy"
author: "dag.tanneberg@wzb.eu" 
date: "December 17, 2015" 
output:
  beamer_presentation: 
    theme: "Berlin"
    fonttheme: "professionalfonts"
    colortheme: "seagull"
    toc: true 
    includes: 
      in_header: header.tex
---

# Why we should talk about this

## Is this the latest thing?

"In an age of widespread secondary school and university 
education, the pervasiveness of the mass media, and the 
displacement of manual labor by clerical and professional 
employees, [the rise of value-oriented intellectuals; D.T.] 
constitutes a challenge to democratic government which is 
[...] as serious as those posed in the past by 
aristocratic cliques, fascist movements, and communist 
parties."

\scriptsize{Michel Crozier, Samuel P. Huntington, and Joji 
Watanuki (1975): \textit{The Crisis of Democracy}, New York 
University Press: New York, p. 7.}

## Some often raised concerns

1. Functional failure of democracy 
    - legalized torture and extra-judicial killings 
2. Negative externalities of economic markets
    - socioeconomic translates into political inequality
3. Rise of anti-democratic forces
    - right-wing populism in Europe and the US

# Negotiating terms

## Balancing democracy and crisis

\begin{figure}[h] \centering 
\includegraphics[width=.8\textwidth]{./termsCorrelation.pdf} 
\end{figure}

## Lessons learned on terminology

**Content conditions scope**

The more things go into democracy the more observations may
qualify as crises.

**Scope changes texture**

As the domain of democracy increases, the range of crisis
shifts from acute, short-term disruptionx to latent,
long-term decline.

# Physical integrity violations in established democracies

```{r workspaceSetup, echo = FALSE, message = FALSE}
rm(list = ls())
packs <- c('ggplot2', 'car', 'lme4')
invisible(lapply(packs, library, character.only = TRUE))
pathOut <- "/Users/dag/Dropbox/publikationen/2013/PIinDemocracy/PiDemProj/output"
load(file.path(pathOut, "base.RData"))
base <- subset(base, year >= 1995 & country != "Malta")

row.names(base) <- with(base,    # creates human readable id
  paste(iso2c, year, sep = ":")
)
base <- within(base, {
  t <- ave(
    year, cow, FUN = function(x) {seq(along.with=x)-1}
  )
  physint.round <- log(physint+1)
  threat.peace <- log(demonstr + strike + 1)
  threat.vio <- log(all.Terror + riot + 1)
  war <- ifelse(interwar == 1 | extrawar == 1, 1, 0)
  std.threat.peace <- scale(threat.peace, scale = FALSE)
  std.threat.vio <- scale(threat.vio, scale = FALSE)
  std.solt.ginmar <- scale(solt.ginmar, scale = FALSE)
  std.gr.pop <- scale(gr.pop, scale = FALSE)
  std.solt.ginet <- scale(solt.ginet, scale = FALSE)
  std.wdi.gini <- scale(wdi.gini, scale = FALSE)
  std.war <- scale(war, scale = FALSE)
  m.ruleoflaw <- ave(ruleoflaw, cow, FUN = mean)/100      ## RoL
  std.ruleoflaw <- scale(m.ruleoflaw, scale = FALSE)
  m.compet <- ave(compet, cow, FUN = mean)/100 ## Accountability
  std.compet <- scale(m.compet, scale = FALSE)
  }
)
```

## Research Questions

1. Do established democracies violate human rights?
2. Why do established democracies violate human rights?
3. What institutions prevent violent government behavior?

## Embedded Democracy at a Glance

\begin{figure}[h] \centering 
\includegraphics[width=.8\textwidth]{./EmbeddedDemocracy.pdf} 
\end{figure}

## Established democracies violate physical integrity.
```{r heatmap, echo = FALSE, fig.align = 'center', fig.width = 4, fig.height = 4/1.5}
# === Heatmap for physint ==================================
df <- subset(                                  ## data setup
  base, select = c("country", "cow", "year", "physint")
)
df <- within(df,       ## compute mean physint and sort data
  mean.physint <- ave(physint, cow, 
    FUN = function(x) {mean(x, na.rm = TRUE) }
  )
)
df <- df[with(df, order(-mean.physint, country)), ]
df <- within(df,         # factor countries for sorting plot
  f.country <- factor(country)
)

library('ggplot2')
library('grid')        ## for unit() function in plot.margin
base_size <- 9                             ## set font size
p <- ggplot(df, aes(
  x = year, 
  y = reorder(f.country, mean.physint, decrease = TRUE))
) +                                            ## add geoms
geom_tile(
  data = df, aes(fill = factor(physint)), colour = "white",
  size = .3
) +
scale_x_continuous(
  breaks = seq(1995, 2007, 3), expand = c(0, 0)
) +
scale_y_discrete(expand = c(0, 0)) +
scale_fill_brewer(type = 'seq', palette = 9) +
theme_minimal(base_size = base_size) +             ## add theme
labs(x = "", y = "", fill = "PI Violations") +
guides(     ## remove white slashes from fill legend boxes
  fill = guide_legend(override.aes = list(colour = NULL))
) +
theme(                      ## finishing touches to theme
  text = element_text(family = 'sans'),
  legend.position = c(.325, 1.04),
  legend.direction = "horizontal",
  legend.key = element_rect(colour = "white"),
  legend.key.size = unit(.5, "lines"),
  panel.background = element_blank(),
  panel.border = element_blank(), 
  panel.grid = element_blank(),
  axis.ticks = element_blank(),
  axis.title = element_blank(),
  axis.text = element_text(size = .6*base_size, angle = 0),
  plot.margin = unit(c(1, .1, 0, 0), "lines") ## crop margins
)
print(p)
rm(df, p)
```

## Horizontal accountability does not constrain.
```{r interactionTerm, echo = FALSE, fig.align = 'center', fig.width = 4, fig.height = 4/1.618, results = FALSE}

library('lme4')
lmer.fit10 <- lmer(
  physint.round ~ 1 + t + std.threat.vio + std.threat.peace +
  std.solt.ginmar + std.gr.pop + war +
  std.ruleoflaw + std.compet + (1 | cow) + (0 + t | cow),
  data = base,
  REML = FALSE
)
lmer.fit11 <- update(lmer.fit10,
  . ~ . + std.threat.vio:std.ruleoflaw,
  data = base,
  REML = FALSE
)
cmult <- 1.96

pred.dta <- expand.grid(
  physint.round = 0,
  std.threat.peace = 0,
  std.threat.vio = base[, 'std.threat.vio'],
  std.solt.ginmar = 0,
  std.gr.pop = 0,
  std.gr.pop = 0,
  war = 0,
  t = 0,
  std.ruleoflaw = quantile(
    unique(base[, 'std.ruleoflaw']),
    probs = c(.05, .25, .5, .75, .95)
  ),
  std.compet = 0
)

## create confidence intervals -----------------------------
mm <- model.matrix(
  terms(lmer.fit11), pred.dta
)
pred.dta[, 'yhat'] <- predict(     # no RE b/c threat fixed
  lmer.fit11, pred.dta, re.form=NA
)
pvar <- diag(mm %*% tcrossprod(vcov(lmer.fit11), mm))
pred.dta <- data.frame(
  pred.dta,
  plo = pred.dta[, 'yhat'] - cmult * sqrt(pvar),
  phi = pred.dta[, 'yhat'] + cmult * sqrt(pvar)
)

## transform data to original scales -----------------------
quantiles.rol <- quantile(         # Define linetype legend
  unique(base[, 'std.ruleoflaw']), 
  probs = c(.05, .25, .5, .75, .95)
)
tmp <- aggregate(m.ruleoflaw ~ country, FUN = mean, data = base)
tmp <- tmp[with(tmp, order(m.ruleoflaw)), ]

q <- 0
pred.dta[, 'quantiles.rol'] <- 0
for(i in quantiles.rol){
  q <- q+1
  pred.dta[, 'quantiles.rol'] <- ifelse(
    pred.dta[, 'std.ruleoflaw'] == i, q, pred.dta[, 'quantiles.rol']
  )
}
rm(i, q)

pred.dta <- within(pred.dta, {
  threat <- std.threat.vio + mean(base$threat.vio, na.rm = TRUE)
  threat <- exp(threat)-1
  rol <- std.ruleoflaw + mean(base$m.ruleoflaw, na.rm = TRUE)
  yhat <- exp(yhat)-1
  plo <- exp(plo)-1
  phi <- exp(phi)-1
  type = 'Violent'
  quantiles.rol <- factor(
    quantiles.rol, levels = 1:5, 
    labels = c(
      '5%', '25%', '50%', '75%', '95%'
    )
  )
  }
)

# Proceed to plotting --------------------------------------
p <- ggplot(
  data = pred.dta,
  aes(y = yhat, x = threat, linetype = quantiles.rol)
) +
geom_line() +
geom_rug(sides = 'b') +
labs(
  x = 'No. of violent political Threats',
  y = 'Fitted PI Violations',
  linetype = 'Rule of Law Quantile'
) +
scale_linetype_manual(
  values = c('solid', 'dotdash', 'dotted', 'longdash', 'twodash')
) +
theme_minimal(base_size = 10) +
theme(
  text = element_text(family = 'sans'),
  legend.position = c(.475, 1.05), 
  legend.direction = 'horizontal',
  legend.key.size = grid::unit(.6, units = 'lines'),
  plot.margin = grid::unit(c(1, 0, 0, 0) + .1, units = 'lines')
)
print(p)
rm(p)
```

# Wrap-up

## Wrap-up

1. The crisis of democracy has been a topic for decades, and
  it touches on real world concerns.
2. What we call democracy changes systematically what we
  call crisis.
3. There may be grain of truth to the debate as established
  democracies do not live up to their own standards.
