---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\maketitle

## Before we start

1. Please, run the following code.
```{r, eval = FALSE}
install.packages(c("tidyverse", "gapminder"))
for(p in c("tidyverse", "gapminder")) {
  library(p, character.only = TRUE)
}
```
2. I could not replicate the data$[abbreviated var name] issue.
3. Mac users with a locale warning on startup: Go to [Stack Overflow](https://stackoverflow.com/questions/9689104/installing-r-on-mac-warning-messages-setting-lc-ctype-failed-using-c) or [rstudio.com](https://community.rstudio.com/t/strange-locale-problems-in-r-after-update-to-mojave/15533/4)


# Introduction

## A Taste of ggplot2

\begin{figure}
  \centering
  \includegraphics[scale = 0.19]{./fig/ch-07-election-county-share.png}
  \caption{Vote Share in the 2012 Presidential Elections by 
    County\footnote{\tiny
      Healan, K. forthc. Data Visualization. A Practical
      Introduction. \href{https://socviz.co/maps.html}{https://socviz.co/maps.html}
      (last access: 10/03/2018).
    }
  }
\end{figure}

## More Tasting: London Bicycle Routes[^99]

\begin{figure}
  \centering
  \includegraphics[scale = .19]{./fig/bike_ggplot.png}
\end{figure}

[^99]: Cheshire, J. 2012. Great maps with ggplot2. \href{http://spatial.ly/2012/02/great-maps-ggplot2/}{http://spatial.ly/2012/02/great-maps-ggplot2/} (last access: 10/06/2018).

## Even More Tasting: Hierarchical Edge Bundling[^98]

\begin{figure}
  \centering
  \includegraphics[scale = .3]{./fig/hierarchical_edge_bundling.png}
\end{figure}

[^98]: Holtz, Y. 2017. The R Graph Gallery.  \href{https://bit.ly/2QKdhLU}{https://bit.ly/2QKdhLU} (last access: 10/06/2018).

## An Embarassing Riches of Options[^97]

\begin{figure}
  \centering
  \includegraphics[width = .9\textwidth]{./fig/flowing_data.png}
\end{figure}

[^97]: Yau, N. 2016. 5 Tips for Learning to Code for Vizualization. \href{https://bit.ly/2yu0fe0}{https://bit.ly/2yu0fe0} (last access: 10/11/2018).

## We will create this modest graph together.

```{r, fig.align='center', echo = FALSE, fig.width = 4, fig.height = 2}
rm(list = ls())
library("ggplot2")
library("gapminder")
p <- ggplot(
  data = gapminder,
  mapping = aes(
    x = reorder(continent, gdpPercap, FUN = median),
    y = gdpPercap
  )
) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar) + # log10 xvalues, but display Dollar
  labs(
    x = "",
    y = "GDP Per Capita",
    title ="Economic Development",
    caption = "Source: Gapminder.",
    shape = "Continent"
  )
p + theme_minimal()
```

## Outline

\tableofcontents

# What is ggplot2?

## Graphics Engines in $\mathcal{R}$

- Three major graphics systems in $\mathcal{R}$
  - base graphics: ships with $\mathcal{R}$[^1]
  - lattice developed by \href{https://www.isid.ac.in/~deepayan/}{Deepayan Sarkar}
  - ggplot2 developed by \href{http://hadley.nz/}{Hadley Wickham}
- All build on Paul Murrell's \href{https://stat.ethz.ch/R-manual/R-patched/library/grid/doc/grid.pdf}{grid Graphics}
- All differ remarkbly on usability & quality of output

[^1]: Check out Nathan Yau's tutorials at \href{https://flowingdata.com/category/tutorials/}{Flowing Data}.

## What's the buzz about ggplot2?

- Implements the Grammar of Graphics[^2]

    "In brief, the grammar tells us that a statistical graphic is a
    mapping from data to aesthetic attributes (colour, shape, size)
    of geometric objects (points, lines, bars)."[^3]

- Very generic, schematic approach to data viz
- Flexible themes & fast, high quality results
- Many users, active community, numerous extension
    - Create animations, networks, maps, dags, etc.


[^2]: Wilkinson, L. 1999/2005. The Grammar of Graphics.  
New York: Springer Science + Business.
[^3]: Wickham, H. 2009. ggplot2. New York: Springer Science + Business.

# How do you talk to ggplot2?

## Intuition

- Idea: logical connection b/w data (e.g., gender) and plot elements (e.g., shapes)
    - a.k.a. aesthetic mapping
- Answer the following questions for ggplot2:
  1. What is your data?
  2. What relationships you want to see?
  3. How do you want to see it?
  4. What additional information do you want to see?
  5. What scales, axes, labels should be swown?
- Complete reference:
  - Hadley's book: [ggplot2](https://link.springer.com/book/10.1007/978-0-387-98141-3)
  - ggplot2 Online Reference: [tidyverse](https://ggplot2.tidyverse.org/reference/)
  - Kieran Healy's book: [Data Vizualization](http://socviz.co/)

## A schematic for making a plot

```{r, eval = FALSE}
p <- ggplot(
	data = <data>, # 1. What is your ("tidy") data?
	mapping = aes( # 2. What var's map unto the plot?
		<aesthetic> = <variable>, # x
		<aesthetic> = <variable>, # y
		<...> = <...> # colour, fill, shape, size,
		  # alpha, ...
	)
) + # Add layers to your plot
  geom_<type>(<...>) + # Define your plot type
  scale_<mapping>_<type>(<...>) +  # Adjust scales
  coord_<type>(<...>) + # Adjust co-ordinates
  labs(<...>) # Label plot elements
ggsave(file = <...>, plot = <...>, ...) # Save ur work
```

## Our target graph
```{r, echo = FALSE, fig.align='center', echo = FALSE, fig.width = 4, fig.height = 2}
rm(list = ls())
library("ggplot2")
library("gapminder")
p <- ggplot(
  data = gapminder,
  mapping = aes(
    x = reorder(continent, gdpPercap, FUN = median),
    y = gdpPercap
  )
) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar) + # log10 xvalues, but display Dollar
  labs(
    x = "",
    y = "GDP Per Capita",
    title ="Economic Development",
    caption = "Source: Gapminder.",
    shape = "Continent"
  )
p + theme_minimal()
```

- Install and load packages ggplot2 and gapminder.
- Answer questions 1-5 on paper.
- Use the generic schematic and \href{https://ggplot2.tidyverse.org/}{tidyverse}
  to generate the plot.

# Summary

## Summary

- There are three major graphics engines: base, ggplot, lattice.
- ggplot2's attraction
    - very generic, schematic approach to viz
    - fast, visually pleasing results
- Principle strategie:
    a. Connect (map) data to plot elements
    b. Layer plot elements
- Read Healan, K. forthc. Data Visualization. A Practical
      Introduction. 
      \href{https://socviz.co/maps.html}{https://socviz.co/maps.html}
      (last access: 10/03/2018).

## Self-practice: Replicate these figures.
```{r ggplot2_self, fig.align='center', echo = FALSE, fig.width = 4.5, fig.height = 2.5}
p <- ggplot(
  data = gapminder,
  mapping = aes(x = gdpPercap, y = lifeExp)
) +
  geom_point(alpha = .2) +
  geom_smooth(method = 'lm', colour = "red", size = .3) + # add linear correlation
  scale_x_log10(labels = scales::dollar) + # log10 xvalues, but display Dollar
  labs(
    x = "GDP per capita",
    y = "Life expectancy at birth",
    title ="Life Expectancy vs. Economic Development"
  ) +
  facet_wrap(~ continent, nrow = 2) + 
  theme_gray(base_size = 9)
p
```

## Self-practice: Replicate this figure.
```{r ggplot2_self2, fig.align='center', echo = FALSE, fig.width = 4, fig.height = 2.7}
p <- ggplot(
  data = gapminder[gapminder$year >= 1997, ],
  mapping = aes(x = gdpPercap, y = lifeExp, colour = continent)
) +
  geom_point(alpha = .6) +
  geom_smooth(method = "loess", aes(colour = NULL)) +
  scale_x_log10(labels = scales::dollar)  +
  labs(
    x = "GDP per capita",
    y = "Life expectancy at birth",
    colour = "Continent",
    title = "Life Expactancy vs. Economic Development",
    subtitle = "Graph shows 1997, 2003, and 2007."
  ) +
  theme_gray(base_size = 9)
p
```
