---
output:
  beamer_presentation:
    theme: Berlin
    colortheme: seagull
    fonttheme: professionalfonts
    highlight: tango
    toc: false
    includes:
      in_header: preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

\maketitle

# Introduction

## A Taste of ggplot2

\begin{figure}
  \centering
  \caption{Vote Share in the 2012 Presidential Elections by 
    County\footnote{\tiny
      Healan, K. forthc. Data Visualization. A Practical
      Introduction. \href{https://socviz.co/maps.html}{https://socviz.co/maps.html} (last access: 10/03/2018).
    }
  }
  \includegraphics[scale = 0.19]{./fig/ch-07-election-county-share.png}
\end{figure}

## More Tasting: Trajectories in Political Repression\footnote{\tiny
      Tanneberg, D. 2018. The Politics of Repression under
      Authoritarian Rule. How Steadfast is the Iron Throne?
      Dissertation Manuscript, Humboldt University Berlin, p. 54.
    }

\begin{figure}
  \centering
  \includegraphics[scale = .43]{./fig/point_ctryTrajectory.pdf}
\end{figure}

## Outline

\tableofcontents

# What is ggplot2?

## Graphics Engines in $\mathcal{R}$

## Why use ggplot2?

# How to talk to ggplot2?

## Our target graph

```{r ggplot2_togeher, fig.align='center', echo = FALSE, fig.width = 4, fig.height = 2.8}
rm(list = ls())
library("ggplot2")
library("gapminder")
p <- ggplot(
  data = gapminder,
  mapping = aes(x = reorder(continent, gdpPercap, FUN = median), y = gdpPercap)
) +
  geom_boxplot() +
  scale_y_log10(labels = scales::dollar) + # log10 xvalues, but display Dollar
  labs(
    x = "",
    y = "GDP Per Capita",
    title ="Economic Development by Continent",
    subtitle = "Boxplots summarize country-years",
    caption = "Source: Gapminder.",
    shape = "Continent"
  )
p
```

# Practice with ggplot2

## Replicate this figure
```{r ggplot2_self, fig.align='center', echo = FALSE, fig.width = 4.5, fig.height = 2.5}
p <- ggplot(
  data = gapminder,
  mapping = aes(x = gdpPercap, y = lifeExp)
) +
  geom_point(alpha = .2) +
  geom_smooth(method = 'lm') + # add linear correlation
  scale_x_log10(labels = scales::dollar) + # log10 xvalues, but display Dollar
  labs(
    x = "",
    y = "",
    title ="Life Expectancy vs. Economic Development"
  ) +
  facet_wrap(~ continent, nrow = 2) + 
  theme_gray(base_size = 9)
p
```
